%!TEX root = ../GeneralRelativity.tex
% ──────────────────────────────────────────────────────────────
%  Lecture 5 — Flows and Tensors
% ──────────────────────────────────────────────────────────────

\section{Flows and tensors}\label{sec:flows-tensors}

In the previous lecture we defined the tangent space $V_p$ at a
point $p\in\M$, coordinate bases, curves, and tangent vector
fields.  We now study how vector fields generate \emph{flows}
(one-parameter groups of diffeomorphisms), how to form new vector
fields from old ones via the \emph{commutator}, and the algebraic
machinery of \emph{tensors} that lies at the heart of general
relativity.

% ──────────────────────────────────────────────────────────────
\subsection{Flows (one-parameter groups of diffeomorphisms)}%
\label{sec:flows}

\begin{definition}[One-parameter group of diffeomorphisms]%
\label{def:one-param-group}
  Let $\M$ be a smooth manifold.  A \textbf{one-parameter group of
  diffeomorphisms} is a $C^\infty$ map
  \[
    \Phi\colon \R\times\M \to \M\,,
    \qquad (t,p)\mapsto \Phi_t(p)\,,
  \]
  such that
  \begin{enumerate}
    \item for each fixed $t\in\R$, the map
      $\Phi_t\colon\M\to\M$ is a diffeomorphism, and
    \item for all $s,t\in\R$,
      \begin{equation}\label{eq:group-law}
        \eqbox{\Phi_s \circ \Phi_t = \Phi_{t+s}}\,.
      \end{equation}
  \end{enumerate}
\end{definition}

\noindent
Choosing $s = -t$ in~\eqref{eq:group-law} immediately gives
$\Phi_0 = \id$ (the identity map on $\M$).

\begin{center}
\begin{tikzpicture}[scale=0.85]
  % Manifold
  \draw[thick, spacecadet, rounded corners=14pt]
    (0,0.2) to[out=30,in=160] (4.5,0.6)
    to[out=-20,in=60] (5.0,-0.8)
    to[out=240,in=-10] (0.5,-1.0)
    to[out=170,in=260] cycle;
  \node[font=\sf\small, text=cgblue] at (4.8,0.8) {$\M$};
  % Point p
  \node[circle, fill=banana, inner sep=1.8pt] (P) at (1.2,-0.1) {};
  \node[font=\small, below left] at (P) {$p$};
  % Flow curve
  \draw[thick, munsell] (P.center)
    to[out=20,in=200] (2.5,0.3)
    to[out=20,in=190] (3.8,0.2);
  % Points along flow
  \node[circle, fill=banana, inner sep=1.5pt] (Q1) at (2.5,0.3) {};
  \node[font=\scriptsize, above] at (Q1) {$\Phi_t(p)$};
  \node[circle, fill=banana, inner sep=1.5pt] (Q2) at (3.8,0.2) {};
  \node[font=\scriptsize, above] at (Q2) {$\Phi_{t+s}(p)$};
  % Tangent vector at p
  \draw[vecstyle] (P) -- ++(0.7,0.25)
    node[above, font=\small]{$v|_p$};
\end{tikzpicture}
\end{center}

\subsubsection{From flows to vector fields}

Given a one-parameter group $\Phi_t$, we obtain a tangent vector
field $v$ on $\M$ as follows.  For a fixed $p\in\M$, the map
\[
  C_p\colon \R\to\M\,,\qquad t\mapsto \Phi_t(p)\,,
\]
is a smooth curve through $p$ (with $C_p(0)=p$), called the
\textbf{orbit} of $p$ under $\Phi_t$.  Define $v|_p$ to
be the tangent vector to $C_p$ at $t=0$:
\begin{equation}\label{eq:flow-to-field}
  v|_p(f) = \td{}{t}(f\circ\Phi_t(p))\bigg|_{t=0}\,,
  \qquad f\in\mathscr{F}(\M)\,.
\end{equation}
Repeating this for every $p\in\M$ yields a smooth vector field $v$.

\subsubsection{From vector fields to flows (integral curves)}

Conversely, given a smooth vector field $v$ on $\M$, we seek
\textbf{integral curves} of $v$: a family of curves in $\M$, one
through each point $p$, such that the tangent vector to the curve
at $p$ equals $v|_p$.

Let $C$ be such an integral curve through $p$, and let $(x^1,\dots,x^n)$
be local coordinates.  Then the tangent vector to $C$ acts on
$f\in\mathscr{F}(\M)$ by
\begin{equation}\label{eq:tangent-to-C}
  T(f) = \td{}{t}(f\circ C)
       = \sum_{\mu=1}^{n} \td{x^\mu}{t}\, X_\mu(f)\,,
\end{equation}
where $X_\mu = \partial/\partial x^\mu|_p$ is the coordinate basis.
Meanwhile, the vector field gives
\begin{equation}\label{eq:field-at-p}
  v(f) = \sum_{\mu=1}^{n} v^\mu\, X_\mu(f)\,.
\end{equation}
Setting~\eqref{eq:tangent-to-C} equal to~\eqref{eq:field-at-p},
we must solve the system of coupled ordinary differential equations
\begin{equation}\label{eq:integral-curve-ode}
  \eqbox{\td{x^\mu}{t} = v^\mu(x^1,\dots,x^n)}
  \qquad \mu = 1,\dots,n\,.
\end{equation}

By the standard existence and uniqueness theorem for ODEs
(proved via fixed-point arguments; see, e.g.,
Coddington and Levinson~\cite{CL55}), there
exists $\varepsilon > 0$ such that a unique solution exists for all
$t\in(-\varepsilon,\varepsilon)$.

\begin{remark}
  If $\M$ is compact, the integral curves exist for all
  $t\in\R$, and the family of integral curves defines a genuine
  one-parameter group of diffeomorphisms.
\end{remark}

\begin{intuition}[Flows as fluid motion]
  Think of $v$ as a velocity field describing steady fluid flow on
  $\M$.  The integral curves are the streamlines: each fluid
  particle follows the integral curve through its initial position.
  The map $\Phi_t$ advances every particle by time~$t$.
\end{intuition}

With flows and integral curves in hand, we have assembled the
basic toolkit for describing ``mechanics on manifolds''---how
points move under smooth evolution.  Notice, however, that nothing
in our discussion so far has involved special or general
relativity: we are still working with bare manifolds, without any
notion of distance, angle, or causal structure.  The machinery we
develop next---the commutator of vector fields, and especially the
algebraic theory of tensors---will supply precisely the additional
structure needed to formulate physics on curved spacetimes.

% ──────────────────────────────────────────────────────────────
\subsection{New vector fields from old: the commutator}%
\label{sec:commutator}

\begin{definition}[Commutator / Lie bracket]\label{def:commutator}
  Let $v$ and $w$ be smooth vector fields on a manifold $\M$.
  Define a new map $u\colon\mathscr{F}(\M)\to\mathscr{F}(\M)$ by
  \begin{equation}\label{eq:commutator}
    \eqbox{u(f) = v\bigl(w(f)\bigr) - w\bigl(v(f)\bigr)}
    \qquad\text{for all } f\in\mathscr{F}(\M)\,.
  \end{equation}
  This map $u$ is called the \textbf{commutator} (or \textbf{Lie
  bracket}) of $v$ and $w$, and is written $u = [v,w]$.
\end{definition}

\begin{exercise}\label{ex:commutator-vf}
  Show that $[v,w]$ as defined in~\eqref{eq:commutator} is
  indeed a vector field, i.e.\ it satisfies linearity and the
  Leibniz rule at every point.
\end{exercise}

\begin{remark}
  Although $v$ and $w$ each involve first-order derivatives, the
  second-order terms cancel in the combination $v(w(f))-w(v(f))$,
  leaving $[v,w]$ a first-order operator---hence a genuine vector
  field.  The commutator plays a central role in the theory of Lie
  groups and will reappear when we study Lie derivatives and the
  curvature of connections.
\end{remark}

% ──────────────────────────────────────────────────────────────
\subsection{New vector spaces from old}\label{sec:new-spaces}

Before defining tensors on manifolds, we review the key algebraic
constructions that build new vector spaces from given ones.  The
guiding observation is this: our manifold $\M$ canonically provides
a single ``atomic'' vector space at each point---the tangent space
$V_p$.  Every other geometric object we need (the metric, the
curvature tensor, differential forms, \ldots) lives not in $V_p$
itself but in some vector space \emph{built from} $V_p$ via
elementary algebraic operations.

One should assess the value of each construction by whether it
produces quantities with a clear physical interpretation.  Just
because an operation is available does not mean physics requires it.

Let $V$ and $W$ be finite-dimensional real vector spaces.  The
principal operations are:

\begin{enumerate}[(i)]
  \item \textbf{Taking duals:} $V \mapsto V^*$.
  \item \textbf{Taking direct sums:} $V,\, W \mapsto V\oplus W$.
  \item \textbf{Taking tensor products:} $V,\, W \mapsto V\tp W$.
  \item \textbf{Taking subspaces:} $U\subset V$.
  \item \textbf{Taking exterior (wedge) products:}
    $V,\, W \mapsto V\wedge W$.
  \item \textbf{Taking joins:} $V,\, W \mapsto V\vee W$.
\end{enumerate}

\noindent
In general relativity, \emph{duals} and \emph{tensor products} are
by far the most important: the metric, for instance, naturally
lives in a tensor product of two copies of $V_p^*$.  Direct sums
play a role in fiber-bundle theory but rarely appear in our
treatment; wedge products feature in the theory of differential
forms; joins and intersections are more prominent in quantum
information theory.  We develop the first three in detail below.

% ──────────────────────────────────────────────────────────────
\subsection{The dual space $V^*$}\label{sec:dual-space}

\begin{definition}[Dual space]\label{def:dual-space}
  Let $V$ be a real vector space.  The \textbf{dual space} $V^*$ is
  the vector space of all linear functions $f\colon V\to\R$.
  Elements of $V^*$ are called \textbf{dual vectors} (or
  \textbf{covectors}, or \textbf{one-forms}).
\end{definition}

\begin{exercise}\label{ex:dual-vs}
  Show that $V^*$ is a real vector space under pointwise addition
  and scalar multiplication.
\end{exercise}

\begin{remark}
  Let $\{v_1,\dots,v_n\}$ be a basis for $V$.  We obtain a
  \textbf{dual basis} $\{(v^\mu)^* \mid \mu=1,\dots,n\}$ for
  $V^*$ defined by
  \begin{equation}\label{eq:dual-basis}
    (v^\mu)^*(v_\nu) = \delta^\mu{}_\nu\,,
  \end{equation}
  where $\delta^\mu{}_\nu$ is the Kronecker delta.
\end{remark}

\noindent
Notice the placement of indices: elements of $V$ carry a subscript
(lower index), while elements of $V^*$ carry a superscript (upper
index) in their labelling.  This convention will become essential
when we work with tensor components on manifolds.

\begin{exercise}\label{ex:double-dual}
  Show that $(V^*)^* \cong V$ (the double dual is canonically
  isomorphic to $V$).  This result will be used frequently and
  without further comment throughout the course.
\end{exercise}

\begin{exercise}\label{ex:dim-dual}
  Show that $\dim(V^*) = \dim(V)$.
\end{exercise}

\begin{remark}
  The map $v_\mu \mapsto (v^\mu)^*$ gives an isomorphism between
  $V$ and $V^*$, but this isomorphism \textbf{depends on the choice
  of basis}---it is \emph{non-canonical}.  Had we chosen a
  different basis for $V$, we would obtain a different
  identification between $V$ and $V^*$.  In physics we seek
  statements that are independent of any particular coordinate
  system chosen by the observer.  A \emph{canonical} (basis-free)
  isomorphism $V_p \to V_p^*$ arises only once we equip the
  manifold with a metric; this is the operation of ``lowering an
  index'' that we shall meet shortly.
\end{remark}

% ──────────────────────────────────────────────────────────────
\subsection{The tensor product}\label{sec:tensor-product}

The tensor product is one of the most versatile constructions in
all of mathematical physics.  Beyond its central role in general
relativity, it appears whenever one deals with multi-particle
systems: the state space of $N$~quantum-mechanical particles is
the tensor product of the individual Hilbert spaces, and tensor
products of qubit spaces underpin quantum information theory and
quantum computing.  We give three perspectives on the
construction---an informal definition, a formal quotient-space
definition, and the dual (multilinear-map) characterisation used in
most relativity texts---so that you can make reference to whichever
viewpoint is most convenient in your future studies.

\subsubsection{Informal definition}

\begin{definition}[Tensor product (informal)]\label{def:tp-informal}
  Let $V$ and $W$ be finite-dimensional real vector spaces.  The
  \textbf{tensor product} $V\tp W$ consists of elements of the
  form $v\tp w$ (for $v\in V$, $w\in W$) together with all their
  linear combinations, subject to the \textbf{bilinearity condition}:
  \begin{equation}\label{eq:bilinearity}
    (a_1 v_1 + a_2 v_2)\tp(b_1 w_1 + b_2 w_2)
    = a_1 b_1\, v_1\tp w_1
    + a_1 b_2\, v_1\tp w_2
    + a_2 b_1\, v_2\tp w_1
    + a_2 b_2\, v_2\tp w_2\,.
  \end{equation}
\end{definition}

\begin{example}\label{ex:R2-tensor-R2}
  Let $V = \Rn{2}$, $W = \Rn{2}$.  Then
  $V\tp W = \Rn{2}\tp\Rn{2} \cong \Rn{4}$.
\end{example}

\subsubsection{Formal construction}

\emph{(This subsection may be skipped on a first reading without
loss of continuity.)}

More precisely, one defines
\begin{equation}\label{eq:tp-formal}
  V\tp W \;\equiv\; F(V\times W)\big/{\sim}\,,
\end{equation}
where $F(V\times W)$ is the free vector space on the set
$V\times W$ (formal linear combinations of ordered pairs $(v,w)$).
Since $V\times W$ is an infinite set, $F(V\times W)$ is
infinite-dimensional and far too large; the equivalence relation
$\sim$ collapses it to the correct finite-dimensional space.  The
relation $\sim$ is generated by:
\begin{enumerate}
  \item identity, symmetry, and transitivity (equivalence relation
    axioms),
  \item distributivity:
    $(v_1+v_2,\,w) \sim (v_1,w) + (v_2,w)$ and
    $(v,\,w_1+w_2) \sim (v,w_1) + (v,w_2)$,
  \item scalar multiples:
    $c(v,w) \sim (cv,\,w) \sim (v,\,cw)$ for all $c\in\R$.
\end{enumerate}
The equivalence class of $(v,w)$ is written $v\tp w$.  Elements of
$V\tp W$ are linear combinations of \textbf{simple tensors}
$v_j\tp w_k$.

\begin{keyresult}[Not every tensor is simple]
  \textbf{Warning:} given $u\in V\tp W$, it is \emph{not} always
  true that $u = v\tp w$ for some $v\in V$, $w\in W$.  Generically,
  $u$ is a sum of simple tensors that cannot be factored.
\end{keyresult}

\begin{exercise}\label{ex:dim-tp}
  Show that $\dim(V\tp W) = \dim(V)\cdot\dim(W)$.
  \emph{Hint:} If $\{v_\mu\}$ is a basis for $V$ and
  $\{w_\nu\}$ is a basis for $W$, show that
  $\{v_\mu\tp w_\nu\}$ is a basis for $V\tp W$.
\end{exercise}

\noindent
The tensor product is associative:
\begin{equation}\label{eq:tp-assoc}
  U\tp(V\tp W) \;\cong\; (U\tp V)\tp W\,,
\end{equation}
so we may write $U\tp V\tp W$ without ambiguity.

\subsubsection{The dual characterisation of tensors}

\begin{definition}[Tensor as multilinear map]\label{def:tensor-multilinear}
  Let $V$ and $W$ be real vector spaces.  A \textbf{tensor} (in
  the dual sense) is a multilinear map
  $T\colon V\times W \to \R$, i.e.\
  \begin{equation}\label{eq:multilinear}
    T(a_1 v_1 + a_2 v_2,\; b_1 w_1 + b_2 w_2)
    = a_1 b_1\, T(v_1,w_1)
    + a_1 b_2\, T(v_1,w_2)
    + a_2 b_1\, T(v_2,w_1)
    + a_2 b_2\, T(v_2,w_2)\,.
  \end{equation}
\end{definition}

\noindent
This is the ``classical'' approach to tensor products, and the one
most commonly seen in general relativity textbooks (including
Wald~\cite{Wald84}).  It is related to the informal definition by
\begin{equation}\label{eq:tp-dual-relation}
  (V\tp W)^* \;\cong\;
  \{T\colon V\times W \to \R \mid T \text{ multilinear}\}\,.
\end{equation}

\begin{exercise}\label{ex:tp-dual-iso}
  Show that $(V\tp W)^* \cong V^*\tp W^*$.
\end{exercise}

% ──────────────────────────────────────────────────────────────
\subsection{Tensors on manifolds}\label{sec:tensors-manifolds}

We now apply these algebraic constructions to the tangent space
$V_p$ at a point $p\in\M$.

From $V_p$ we build:
\begin{itemize}
  \item the dual space $V_p^*$ (cotangent vectors),
  \item tensor products such as $V_p\tp V_p$,
    $V_p^*\tp V_p$, etc.,
  \item and more generally, the space
    $\underbrace{V_p\tp\cdots\tp V_p}_{k}\,\tp\,
     \underbrace{V_p^*\tp\cdots\tp V_p^*}_{l}
     \;\cong\; V_p^{\tp k}\tp (V_p^*)^{\tp l}$.
\end{itemize}

\noindent
Since $(V_p^*)^* \cong V_p$, an element $t\in V_p$ can be viewed
as a linear function on $V_p^*$ as well.

One could also form ``mixed'' tensor products such as
$V_p\tp V_p^*\tp V_p$, where the factors are interleaved.  In
general relativity, however, the order of the individual
$V_p$ and~$V_p^*$ factors rarely matters (many of the tensors we
encounter have symmetry properties that make the ordering
irrelevant), so it is conventional to cluster all $V_p$ factors to
the left and all $V_p^*$ factors to the right.  In quantum
information theory, by contrast, the ordering of tensor-product
factors carries physical significance (each factor corresponds to a
distinct subsystem), and one must keep careful track of it.

\subsubsection{Tensors of type $(k,l)$}

\begin{definition}[Tensor of type $(k,l)$]\label{def:tensor-type}
  A \textbf{tensor of type $(k,l)$} at $p$ is an element
  \[
    T \;\in\; V_p^{\tp k}\tp (V_p^*)^{\tp l}\,.
  \]
  Equivalently, $T$ is a multilinear map
  \[
    T\colon \underbrace{V_p^*\times\cdots\times V_p^*}_{k}
      \;\times\;
      \underbrace{V_p\times\cdots\times V_p}_{l}
      \;\longrightarrow\; \R\,.
  \]
  The vector space of all tensors of type $(k,l)$ at $p$ is denoted
  $\ttype{k}{l}$.
\end{definition}

\begin{equation}\label{eq:dim-tensor-space}
  \dim\bigl(\ttype{k}{l}\bigr) = n^{k+l}\,,
  \qquad\text{where } n = \dim(V_p)\,.
\end{equation}

\begin{remark}
  We adopt the convention $V^{\tp 0} \cong \R$, so that
  $\ttype{0}{0} \cong \R$ (scalars).
\end{remark}

% ──────────────────────────────────────────────────────────────
\subsection{Examples of tensors}\label{sec:tensor-examples}

\begin{example}[Type $(0,1)$: dual vectors]\label{ex:type-01}
  A tensor of type $(0,1)$ is an element of
  $V_p^{\tp 0}\tp (V_p^*)^{\tp 1} \cong \R\tp V_p^* \cong V_p^*$.
  It is simply a \textbf{dual vector} (covector).
\end{example}

\begin{example}[Type $(1,0)$: vectors]\label{ex:type-10}
  A tensor of type $(1,0)$ is an element of
  $V_p^{\tp 1}\tp (V_p^*)^{\tp 0} \cong V_p\tp \R \cong V_p$.
  It is simply a \textbf{vector}.
\end{example}

\begin{example}[Type $(1,1)$: linear maps]\label{ex:type-11}
  A tensor of type $(1,1)$ is an element of $V_p\tp V_p^*$.
  Such a tensor $T$ can be interpreted in three equivalent ways:
  \begin{enumerate}[(i)]
    \item as an element of $V_p\tp V_p^*$,
    \item as a multilinear map $V_p^*\times V_p \to \R$
      (equivalently, a dual vector in $(V_p^*\tp V_p)^*$),
    \item as a \textbf{linear transformation}
      $T\colon V_p\to V_p$, i.e.\ a matrix.
  \end{enumerate}
\end{example}

% ──────────────────────────────────────────────────────────────
\subsection{Tensor components and the action of tensors}%
\label{sec:tensor-components}

Choose a basis $\{v_\mu\}$ of $V_p$ with dual basis
$\{(v^\nu)^*\}$ for $V_p^*$.  A type-$(1,1)$ tensor $T$ can be
expanded as
\begin{equation}\label{eq:type11-expansion}
  T = \sum_{\mu,\nu} t^\mu{}_\nu\; v_\mu \tp (v^\nu)^*
  \;\in\; V_p\tp V_p^*\,.
\end{equation}
The scalars $[T]^\mu{}_\nu = t^\mu{}_\nu$ are called the
\textbf{components} (or \textbf{matrix elements}) of $T$ in the
basis $\{v_\mu\}$.

\medskip
\noindent\textbf{Action on a vector.}\;
Let $w\in V_p$, with $w = \sum_\nu w^\nu\, v_\nu$.  Then
\begin{align}
  Tw &= \biggl(\sum_{\mu,\nu} t^\mu{}_\nu\; v_\mu\tp (v^\nu)^*
        \biggr)(w)
      = \sum_{\mu,\nu} t^\mu{}_\nu\; (v^\nu)^*(w)\; v_\mu
        \notag\\
     &= \sum_{\mu,\nu} t^\mu{}_\nu\; w^\nu\; v_\mu\,.
        \label{eq:tensor-action}
\end{align}
Therefore the components of $Tw$ are
\begin{equation}\label{eq:tensor-action-components}
  \eqbox{[Tw]^\mu = \sum_{\nu=1}^{n} t^\mu{}_\nu\, w^\nu}\,.
\end{equation}
This is an example of \textbf{tensor contraction}.

% ──────────────────────────────────────────────────────────────
\subsection{Tensor contraction}\label{sec:contraction}

\begin{definition}[Contraction]\label{def:contraction}
  Let $T\in\ttype{k}{l}$ with $k\geq 1$ and $l\geq 1$.  The
  \textbf{contraction} over the $j$-th contravariant and $j'$-th
  covariant index is the map
  \[
    C_{j,j'}\colon \ttype{k}{l} \to \ttype{k-1}{l-1}
  \]
  defined (interpreting $T$ as a multilinear map) by
  \begin{equation}\label{eq:contraction}
    (C_{j,j'}\, T)(\dots) =
    \sum_{\sigma=1}^{n}
      T\bigl(\dots,\underbrace{(v^\sigma)^*}_{j\text{-th slot}},
             \dots;\;
             \underbrace{v_\sigma}_{j'\text{-th slot}},\dots\bigr)\,,
  \end{equation}
  where $\{v_\sigma\}$ is any basis for $V_p$ and $\{(v^\sigma)^*\}$
  is the dual basis.
\end{definition}

\noindent
In terms of components:
\begin{equation}\label{eq:contraction-components}
  \eqbox{(C_{j,j'}\, T)^{\mu_1\cdots\mu_{k-1}}{}_{\nu_1\cdots\nu_{l-1}}
  = \sum_{\sigma=1}^{n}
    T^{\mu_1\cdots\sigma\cdots\mu_{k-1}}{}_{\nu_1\cdots\sigma\cdots\nu_{l-1}}}
\end{equation}
where $\sigma$ appears in the $j$-th upper position and the
$j'$-th lower position and is summed over.

\begin{exercise}\label{ex:contraction-basis-ind}
  Show that the contraction is independent of the choice of basis.
\end{exercise}

% ──────────────────────────────────────────────────────────────
\subsection{The outer product}\label{sec:outer-product}

\begin{definition}[Outer product]\label{def:outer-product}
  Let $S\in\ttype{k}{l}$ and $T\in\ttype{k'}{l'}$.  Their
  \textbf{outer product} (or \textbf{tensor product})
  $S\tp T \in \ttype{k+k'}{l+l'}$ is defined in components by
  \begin{equation}\label{eq:outer-product}
    (S\tp T)^{\mu_1\cdots\mu_k\,\mu_{k+1}\cdots\mu_{k+k'}}
             {}_{\nu_1\cdots\nu_l\,\nu_{l+1}\cdots\nu_{l+l'}}
    = S^{\mu_1\cdots\mu_k}{}_{\nu_1\cdots\nu_l}\;\cdot\;
      T^{\mu_{k+1}\cdots\mu_{k+k'}}{}_{\nu_{l+1}\cdots\nu_{l+l'}}\,.
  \end{equation}
\end{definition}

\noindent
Intrinsically, $S\tp T$ lives in
\[
  \bigl(V_p^{\tp k}\tp (V_p^*)^{\tp l}\bigr)
  \tp
  \bigl(V_p^{\tp k'}\tp (V_p^*)^{\tp l'}\bigr)\,,
\]
which is rearranged (collecting all $V_p$ factors to the left and
all $V_p^*$ factors to the right) to give an element of
$V_p^{\tp(k+k')}\tp (V_p^*)^{\tp(l+l')}$.

\begin{intuition}[Building tensors]
  Every tensor can be built from vectors and dual vectors using the
  outer product, and then contracted.  The outer product increases
  the type, while contraction decreases it.  Together, these two
  operations generate the full algebra of tensor manipulations used
  throughout general relativity.
\end{intuition}
