%!TEX root = ../GeneralRelativity.tex
% ──────────────────────────────────────────────────────────────
%  Addendum A — The energy-momentum tensor: a rigorous
%               distributional construction via the coarea formula
% ──────────────────────────────────────────────────────────────

\section{Addendum: the energy-momentum tensor from microscopics}%
\label{sec:emt-rigorous}

In Lecture~6 (\S\ref{sec:energy-momentum}) we defined the
energy-momentum tensor of a system of point particles via the
microscopic expressions~\eqref{eq:Tmunu-dt}
and~\eqref{eq:Tmunu-covariant}, and left the proof that these two
forms are equivalent---and that the result is a Lorentz tensor---as
an exercise (Exercise~\ref{ex:Tmunu-tensor}).  The standard argument,
due to Weinberg~\cite{Weinberg72}, consists of inserting a
coordinate-time integral, observing that ``the differentials~$dt'$
cancel,'' and replacing~$dt'$ with the invariant proper-time
element~$d\tau$.  While physically convincing, this argument glosses
over the distributional nature of the objects involved and leaves the
equivalence of the two forms essentially unproved.

In this addendum we supply the missing rigour.  For the equivalence
theorem itself (\S\ref{sec:equivalence}), the key step is the
classical \textbf{change-of-variables (substitution) formula} for a
$C^1$ diffeomorphism---an elementary result that needs no heavy
machinery.  However, the natural \emph{generalisation} to slicing by
arbitrary Cauchy surfaces (\S\ref{sec:general-slicing}) requires the
\textbf{coarea formula} of geometric measure
theory~\cite{Federer69}, which provides the framework for
decomposing integrals over level sets of Lipschitz maps.
We therefore state the coarea formula in its general form and explain
how the one-dimensional case used in the equivalence proof reduces to
the classical substitution rule, while the full coarea/slicing
machinery becomes essential for the general slicing theorem.

% ──────────────────────────────────────────────────────────────
\subsection{Distributions supported on worldlines}%
\label{sec:dist-worldlines}

Throughout this addendum, $(\Rn{4}, \eta_{\alpha\beta})$ denotes
Minkowski spacetime with signature~$(-,+,+,+)$, and
$\mathcal{D}(\Rn{4}) = C_c^\infty(\Rn{4})$ is the space of
compactly supported smooth test functions.  We write
$\mathcal{D}'(\Rn{4})$ for the topological dual: the space of
distributions (continuous linear functionals on~$\mathcal{D}$).

\begin{definition}[Timelike worldline]\label{def:timelike-worldline}
  A \textbf{timelike worldline} is a smooth, future-directed
  embedding $\gamma\colon I \to \Rn{4}$ (where $I\subseteq\R$ is
  an open interval) parametrised by proper time~$\tau$, so that
  \begin{equation}\label{eq:timelike-condition}
    \eta_{\alpha\beta}\,\dot\gamma^\alpha\dot\gamma^\beta = -1
    \qquad\text{and}\qquad
    \frac{d\gamma^0}{d\tau} > 0
  \end{equation}
  everywhere on~$I$.  Here
  $\dot\gamma^\alpha = d\gamma^\alpha/d\tau$.  Throughout this
  addendum we assume $I = \R$, i.e.\ the worldline is
  inextendible.
\end{definition}

\noindent
For a massive particle of rest mass~$m$, the
\textbf{four-momentum} along~$\gamma$ is
$p^\alpha(\tau) = m\,\dot\gamma^\alpha(\tau)$.

The microscopic energy-momentum tensor is built from delta
functions localised on worldlines.  To make this precise, we define
it directly as a distribution.

\begin{definition}[Distributional energy-momentum tensor]%
\label{def:Tmunu-dist}
  Given $N$ timelike worldlines $\gamma_j\colon \R \to \Rn{4}$
  carrying four-momenta $p_j^\alpha(\tau)$, the
  \textbf{(distributional) energy-momentum tensor}
  $T^{\alpha\beta} \in \mathcal{D}'(\Rn{4})$ is defined by its
  action on test functions:
  \begin{equation}\label{eq:Tmunu-dist-def}
    \eqbox{\langle T^{\alpha\beta},\, \phi\rangle
      \;=\; \sum_{j=1}^{N} \int_{\R} p_j^\alpha(\tau)\,
        \dot\gamma_j^\beta(\tau)\,
        \phi\bigl(\gamma_j(\tau)\bigr)\;\dd\tau}
  \end{equation}
  for every $\phi \in \mathcal{D}(\Rn{4})$.
\end{definition}

\begin{proposition}[Well-definedness]\label{prop:Tmunu-well-defined}
  The functional~\eqref{eq:Tmunu-dist-def} defines a distribution
  of order zero, that is, a tensor-valued Radon measure
  on~$\Rn{4}$.
\end{proposition}

\begin{proof}
  It suffices to treat a single worldline~$\gamma$; the sum
  over~$j$ follows by linearity.

  \medskip\noindent
  \textbf{Step~1: $\gamma^0$ is a diffeomorphism
  $\R\to\R$.}\;
  By Definition~\ref{def:timelike-worldline}, the worldline is
  future-directed:
  $\dot\gamma^0 > 0$.  The normalisation
  $\eta_{\alpha\beta}\,\dot\gamma^\alpha\dot\gamma^\beta = -1$
  with signature $(-,+,+,+)$ gives
  $(\dot\gamma^0)^2 = 1 + \sum_{i=1}^{3}(\dot\gamma^i)^2 \geq 1$,
  so $\dot\gamma^0 \geq 1$.  Hence $\gamma^0(\tau) \geq \gamma^0(0) + \tau$
  for $\tau > 0$ and $\gamma^0(\tau) \leq \gamma^0(0) + \tau$ for
  $\tau < 0$, which forces $\gamma^0 \to \pm\infty$ as
  $\tau\to\pm\infty$.  Thus $f\colon \R\to\R$,
  $f(\tau) = \gamma^0(\tau)$, is a strictly increasing smooth
  surjection and therefore a diffeomorphism $\R\to\R$.

  \medskip\noindent
  \textbf{Step~2: $\gamma$ is proper.}\;
  We must show that $\gamma^{-1}(K)$ is compact for every compact
  $K\subset\Rn{4}$.
  \begin{enumerate}
    \item[(i)] \emph{Bounded $\tau$-range.}\;
      Since $K$ is compact, its time projection
      $\pi_0(K) \subset \R$ is a compact interval.  The
      diffeomorphism~$f$ maps $\gamma^{-1}(K)$ into
      $f^{-1}(\pi_0(K))$, which is a compact interval (as the
      preimage of a compact interval under a diffeomorphism
      $\R\to\R$).  Hence $\gamma^{-1}(K)$ is contained in a
      bounded $\tau$-interval.
    \item[(ii)] \emph{Bounded spatial coordinates.}\;
      The normalisation
      $\eta_{\alpha\beta}\,\dot\gamma^\alpha\dot\gamma^\beta = -1$
      with signature $(-,+,+,+)$ gives
      $\sum_{i=1}^{3}(\dot\gamma^i)^2 = (\dot\gamma^0)^2 - 1
      < (\dot\gamma^0)^2$,
      so $|\dot\gamma^i| < \dot\gamma^0$ for each spatial
      component.  In particular, the spatial components of~$\gamma$
      grow no faster than~$\gamma^0$, so they remain bounded on
      the bounded $\tau$-interval from~(i).
    \item[(iii)] \emph{Compactness.}\;
      Since $\gamma$ is an embedding (hence a homeomorphism onto
      its image), $\gamma^{-1}(K)$ is closed in~$\R$ (as the
      preimage of a compact, hence closed, set under a continuous
      map).  By~(i) and~(ii), it is also bounded.  Hence
      $\gamma^{-1}(K)$ is compact by the Heine--Borel theorem.
  \end{enumerate}

  \medskip\noindent
  \textbf{Step~3: The integral converges and yields
  order-zero continuity.}\;
  Let $\phi \in \mathcal{D}(\Rn{4})$ have
  support in a compact set~$K$.  The function
  $\tau \mapsto p^\alpha(\tau)\,\dot\gamma^\beta(\tau)\,
  \phi(\gamma(\tau))$ is smooth with support in the compact set
  $\gamma^{-1}(K)$, so the integral converges absolutely.
  Linearity of the functional in~$\phi$ is immediate.  Moreover,
  \[
    |\langle T^{\alpha\beta}, \phi\rangle|
      \;\leq\; \sum_j \int_{\gamma_j^{-1}(K)}
        |p_j^\alpha|\,|\dot\gamma_j^\beta|\;\dd\tau
        \cdot \|\phi\|_\infty
      \;\leq\; C_K\,\|\phi\|_\infty\,,
  \]
  where, for fixed $\alpha,\beta$,
  $C_K = \sum_j \int_{\gamma_j^{-1}(K)}
  |p_j^\alpha|\,|\dot\gamma_j^\beta|\,\dd\tau < \infty$ depends
  only on~$K$ (not on~$\phi$).  Thus $T^{\alpha\beta}$ is
  continuous in the $C^0$-topology and hence is a distribution of
  order zero.  (By the Riesz representation theorem, every
  distribution of order zero corresponds to a Radon measure,
  so $T^{\alpha\beta}$ is indeed a tensor-valued Radon measure
  on~$\Rn{4}$.)
\end{proof}

\begin{remark}
  Definition~\ref{def:Tmunu-dist} is the \emph{pushforward} of a
  measure on the worldline to a distribution on spacetime.  In the
  language of de~Rham's theory of currents, the functional
  $\phi \mapsto \int \phi(\gamma(\tau))\,\dd\tau$ defines a
  $0$-current on~$\Rn{4}$ (it acts on $0$-forms, i.e.\ test
  functions).  The \emph{integration current} of the worldline,
  which acts on $1$-forms via
  $\omega \mapsto \int \omega_\alpha(\gamma(\tau))\,
  \dot\gamma^\alpha(\tau)\,\dd\tau$, is a $1$-current.
  Each component $T^{\alpha\beta}$ (for fixed
  $\alpha,\beta$) is obtained from the $0$-current by
  weighting the integrand with
  $p^\alpha\,\dot\gamma^\beta$.  See
  Federer~\cite[\S4.1]{Federer69}.
\end{remark}

\noindent
Informally, Definition~\ref{def:Tmunu-dist} says that
$T^{\alpha\beta}$ is the distribution
\begin{equation}\label{eq:Tmunu-delta4-add}
  T^{\alpha\beta}(x) = \sum_{j=1}^{N} \int\!\dd\tau\;
    p_j^\alpha(\tau)\,\dot\gamma_j^\beta(\tau)\,
    \delta^{(4)}\!\bigl(x - \gamma_j(\tau)\bigr)\,,
\end{equation}
which is precisely equation~\eqref{eq:Tmunu-covariant}.  The
distributional definition makes it clear that this expression is
well defined and coordinate-independent; what remains is to show
that it reduces to the $3{+}1$
form~\eqref{eq:Tmunu-dt} upon slicing by a constant-time
hyperplane.

% ──────────────────────────────────────────────────────────────
\subsection{The coarea formula and the substitution rule}%
\label{sec:coarea}

The bridge between the covariant
form~\eqref{eq:Tmunu-delta4-add} and the $3{+}1$
splitting~\eqref{eq:Tmunu-dt} is ultimately the classical
\emph{change-of-variables formula} for a diffeomorphism.  To place
this in its proper context---and to prepare for the general slicing
theorem of \S\ref{sec:general-slicing}---we state the coarea formula
of Federer and then discuss the one-dimensional case separately.

\begin{theorem}[Coarea formula {\cite[Theorem~3.2.12]{Federer69}}]%
\label{thm:coarea}
  Let $f\colon \Rn{n} \to \Rn{m}$ be a Lipschitz map with
  $m < n$ (strict inequality).  Then for any nonnegative Borel
  function $g\colon \Rn{n} \to [0,\infty]$,
  \begin{equation}\label{eq:coarea-general}
    \int_{\Rn{n}} g(x)\,J_m f(x)\;\dd\mathcal{H}^n(x)
      = \int_{\Rn{m}}
        \biggl(\int_{f^{-1}(y)}
          g\;\dd\mathcal{H}^{n-m}\biggr)
        \dd\mathcal{H}^m(y)\,,
  \end{equation}
  where $J_m f$ denotes the $m$-dimensional Jacobian of~$f$
  and $\mathcal{H}^k$ the $k$-dimensional Hausdorff measure
  (which coincides with the Lebesgue measure on~$\Rn{k}$;
  see~\cite[2.10.35]{Federer69}).
\end{theorem}

\noindent
\textbf{The one-dimensional case ($n = m = 1$).}\enspace
When domain and codomain have the same dimension, the strict
inequality $m < n$ is \emph{not} satisfied, so
Theorem~\ref{thm:coarea} does not apply.  The relevant result is
instead the \textbf{area formula}
\cite[Theorem~3.2.3]{Federer69}: for $f\colon \R \to \R$
Lipschitz and $g\colon \R \to [0,\infty]$ Borel,
\begin{equation}\label{eq:area-1d}
  \eqbox{
  \int_{\R} g(\tau)\,|f'(\tau)|\;\dd\tau
    = \int_{\R}
      \sum_{\tau\in f^{-1}(t)} g(\tau)\;\dd t}\,.
\end{equation}
When $f$ is injective (in particular, a diffeomorphism), each level
set $f^{-1}(t)$ is at most a singleton and the sum is trivial: one
recovers the classical \textbf{substitution (change-of-variables)
formula}
\begin{equation}\label{eq:substitution}
  \int_{\R} g(\tau)\,|f'(\tau)|\;\dd\tau
    = \int_{\R} g\bigl(\tau_\ast(t)\bigr)\;\dd t\,,
\end{equation}
where $\tau_\ast(t) = f^{-1}(t)$ is the unique preimage.  This is
the identity that drives the proof of
Theorem~\ref{thm:equivalence}; the full coarea machinery is not
needed for that argument but becomes essential in
\S\ref{sec:general-slicing}.

\begin{remark}[Delta-function composition rule]%
\label{rem:delta-composition}
  The substitution formula~\eqref{eq:substitution} is equivalent to
  the distributional identity
  \begin{equation}\label{eq:delta-composition}
    \delta\bigl(f(\tau) - t\bigr)
      = \frac{\delta(\tau - \tau_\ast)}{|f'(\tau_\ast)|}\,.
  \end{equation}
  To see this, note that \eqref{eq:substitution} states that
  the pushforward of the measure $|f'(\tau)|\,\dd\tau$ under~$f$
  equals Lebesgue measure~$\dd t$ (by the Riesz representation
  theorem, since the identity holds for all nonneg.\ Borel~$g$).
  Testing against a Dirac mass at~$t$ then
  yields~\eqref{eq:delta-composition}.  More generally, when $f$ has
  finitely many roots $\{\tau_k\}$ of $f(\tau) = t$, the area
  formula~\eqref{eq:area-1d} gives the standard multi-root identity
  $\delta(f(\tau) - t) = \sum_k
  \delta(\tau - \tau_k)/|f'(\tau_k)|$.  The substitution formula
  thus provides the rigorous foundation for the ``delta-function
  composition rule'' that is often invoked without proof in physics
  texts.
\end{remark}

% ──────────────────────────────────────────────────────────────
\subsection{From the covariant form to the spatial-slice form}%
\label{sec:equivalence}

We now prove the main result: the covariant
definition~\eqref{eq:Tmunu-delta4-add} reproduces the $3{+}1$
form~\eqref{eq:Tmunu-dt} of Lecture~6 upon slicing by a
constant-time hyperplane.

\begin{theorem}[Equivalence of forms]\label{thm:equivalence}
  Let $T^{\alpha\beta}$ be the distribution defined
  in~\eqref{eq:Tmunu-dist-def}.  Then, in the sense of
  distributions on~$\Rn{4}$,
  \begin{equation}\label{eq:equivalence}
    T^{\alpha\beta}(x) = \sum_{j=1}^{N} p_j^\alpha(t)\,
      \frac{d\gamma_j^\beta}{dt}(t)\,
      \delta^{(3)}\!\bigl(\vb{x} - \boldsymbol{\gamma}_j(t)\bigr)\,,
  \end{equation}
  where $t = x^0$ and $\boldsymbol{\gamma}_j(t) =
  \bigl(\gamma_j^1(t),\gamma_j^2(t),\gamma_j^3(t)\bigr)$ is the
  spatial position of the $j$-th particle at coordinate time~$t$.
\end{theorem}

\begin{proof}
  It suffices to prove the identity for a single worldline; the
  sum over~$j$ follows by linearity.  Drop the subscript~$j$ and
  let $\gamma\colon \R \to \Rn{4}$ be a timelike worldline
  carrying momentum~$p^\alpha$.

  \medskip\noindent
  \textbf{Step~1: Factor out the time velocity.}\;
  Define the \emph{time function along the worldline}
  $f\colon \R \to \R$ by $f(\tau) = \gamma^0(\tau)$.
  Since $\gamma$ is future-directed and timelike,
  $f'(\tau) = d\gamma^0/d\tau > 0$, so $f$ is a smooth
  diffeomorphism from~$\R$ onto~$\R$.

  Decompose the four-velocity:
  \[
    \dot\gamma^\beta(\tau)
      = \frac{d\gamma^\beta}{d\tau}
      = \frac{d\gamma^\beta}{dt}\bigg|_{t=f(\tau)}\,f'(\tau)\,.
  \]
  Substituting into Definition~\ref{def:Tmunu-dist}:
  \begin{equation}\label{eq:step1}
    \langle T^{\alpha\beta},\phi\rangle
      = \int_{\R} p^\alpha(\tau)\,
        \frac{d\gamma^\beta}{dt}\bigg|_{t=f(\tau)}\,
        \phi\bigl(\gamma(\tau)\bigr)\,f'(\tau)\;\dd\tau\,.
  \end{equation}

  \medskip\noindent
  \textbf{Step~2: Change of variables.}\;
  Since $f$ is a smooth diffeomorphism with $f' > 0$, the level
  set $f^{-1}(t)$ is a singleton $\{\tau_\ast(t)\}$ for each~$t$,
  and the classical change of variables (substitution) gives
  \[
    \int_{\R} g(\tau)\,f'(\tau)\;\dd\tau
      = \int_{\R} g\bigl(f^{-1}(t)\bigr)\;\dd t
  \]
  for any integrable~$g$, without sign restriction on the
  integrand.  (This is the diffeomorphism case of the
  one-dimensional area formula; see
  Remark~\ref{rem:delta-composition}.)  Applying this
  to~\eqref{eq:step1}:
  \begin{equation}\label{eq:step2}
    \langle T^{\alpha\beta},\phi\rangle
      = \int_{\R} p^\alpha\bigl(\tau_\ast(t)\bigr)\,
        \frac{d\gamma^\beta}{dt}\bigg|_{t}\,
        \phi\bigl(\gamma(\tau_\ast(t))\bigr)\;\dd t\,.
  \end{equation}

  \medskip\noindent
  \textbf{Step~3: Identify the spatial delta function.}\;
  By definition of $\tau_\ast(t)$, we have
  $\gamma^0(\tau_\ast(t)) = t$, so
  $\gamma(\tau_\ast(t)) = \bigl(t,\,\boldsymbol{\gamma}(t)\bigr)$.
  Writing $p^\alpha(\tau_\ast(t)) = p^\alpha(t)$ and
  $d\gamma^\beta/dt\,\big|_{\tau_\ast(t)} = d\gamma^\beta/dt\,(t)$
  as functions of coordinate time, equation~\eqref{eq:step2}
  becomes
  \begin{align}
    \langle T^{\alpha\beta},\phi\rangle
      &= \int_{\R} p^\alpha(t)\,
         \frac{d\gamma^\beta}{dt}(t)\,
         \phi\bigl(t,\boldsymbol{\gamma}(t)\bigr)\;\dd t
         \notag\\
      &= \int_{\R}\dd t\int_{\Rn{3}}\dd^3\!\vb{x}\;\;
         p^\alpha(t)\,\frac{d\gamma^\beta}{dt}(t)\,
         \delta^{(3)}\!\bigl(\vb{x} - \boldsymbol{\gamma}(t)\bigr)\,
         \phi(t,\vb{x})\,.
         \label{eq:step3}
  \end{align}
  where the second line uses the sifting property of
  $\delta^{(3)}$ in the reverse direction, and we apply Fubini's
  theorem to pass to the iterated integral (the integrand is
  in $L^1(\R\times\Rn{3})$ since $\phi$ has compact support).
  Since~\eqref{eq:step3} holds for all
  $\phi\in\mathcal{D}(\Rn{4})$, we conclude
  that~\eqref{eq:equivalence} holds in the distributional sense.
\end{proof}

\begin{intuition}[Slicing worldlines by time]
  The theorem says that the natural (covariant) definition of
  $T^{\alpha\beta}$---as a measure spread along the worldline---can
  be ``sliced'' by a constant-time hyperplane $\Sigma_t = \{x^0 = t\}$
  to recover the $3{+}1$ expression.  The change of variables provides
  exactly the right Jacobian factor ($1/f'(\tau_\ast)$) to convert
  the proper-time line element~$\dd\tau$ along the worldline into
  the transverse coordinate~$\dd t$.  This factor cancels the
  $d\gamma^\beta/d\tau$ in the covariant integrand, producing
  $d\gamma^\beta/dt$ in the sliced expression.

  Weinberg's observation that ``the $dt'$ cancel'' is a compact
  (if informal) summary of exactly this mechanism.
\end{intuition}

% ──────────────────────────────────────────────────────────────
\subsection{Lorentz covariance}\label{sec:lorentz-covariance-add}

The covariant form~\eqref{eq:Tmunu-dist-def} immediately yields
the transformation law, resolving
Exercise~\ref{ex:Tmunu-tensor}:

\begin{proposition}[Lorentz covariance]\label{prop:lorentz-cov}
  Under a Lorentz transformation
  $x'^\mu = \Lambda^\mu{}_\nu\, x^\nu$, the distributional
  energy-momentum tensor transforms as
  \begin{equation}\label{eq:Tmunu-lorentz}
    T'^{\alpha\beta}(x')
      = \Lambda^\alpha{}_\gamma\,
        \Lambda^\beta{}_\delta\,
        T^{\gamma\delta}(x)\,.
  \end{equation}
\end{proposition}

\begin{proof}
  Under $x' = \Lambda x$, the transformed worldline is
  $\gamma'(\tau) = \Lambda\gamma(\tau)$ and the transformed
  four-momentum is $p'^\alpha(\tau) = \Lambda^\alpha{}_\gamma\,
  p^\gamma(\tau)$.  (Since $\tau$ is Lorentz-invariant and
  $\gamma^\alpha$ transforms as a four-vector,
  $\dot\gamma^\alpha = d\gamma^\alpha/d\tau$ transforms as a
  four-vector, and hence
  $p^\alpha = m\dot\gamma^\alpha$ does as well.)

  We define the transformed distribution~$T'^{\alpha\beta}$ by
  its action on test functions: for any
  $\phi \in \mathcal{D}(\Rn{4})$,
  \begin{align}
    \langle T'^{\alpha\beta},\,\phi\rangle
      &= \sum_j \int_{\R} p_j'^{\,\alpha}(\tau)\,
        \dot\gamma_j'^{\,\beta}(\tau)\,
        \phi\bigl(\gamma_j'(\tau)\bigr)\;\dd\tau
      \notag\\
      &= \sum_j \int_{\R}
        \Lambda^\alpha{}_\gamma\, p_j^\gamma(\tau)\;\cdot\;
        \Lambda^\beta{}_\delta\, \dot\gamma_j^\delta(\tau)\;\cdot\;
        \phi\bigl(\Lambda\gamma_j(\tau)\bigr)\;\dd\tau
      \notag\\
      &= \Lambda^\alpha{}_\gamma\,\Lambda^\beta{}_\delta
        \sum_j \int_{\R} p_j^\gamma(\tau)\,
        \dot\gamma_j^\delta(\tau)\,
        (\phi \circ \Lambda)\bigl(\gamma_j(\tau)\bigr)\;\dd\tau
      \notag\\
      &= \Lambda^\alpha{}_\gamma\,\Lambda^\beta{}_\delta\;
        \bigl\langle T^{\gamma\delta},\;
        \phi \circ \Lambda\bigr\rangle\,.
      \label{eq:Tmunu-lorentz-dist}
  \end{align}
  This is the standard distributional transformation law
  $T'^{\alpha\beta} = \Lambda^\alpha{}_\gamma\,
  \Lambda^\beta{}_\delta\,(\Lambda^{-1})_\ast T^{\gamma\delta}$,
  which is the distributional restatement
  of~\eqref{eq:Tmunu-lorentz}.
\end{proof}

\begin{remark}
  The $3{+}1$ form~\eqref{eq:Tmunu-dt} is \emph{not} manifestly
  covariant: neither $\delta^{(3)}$ nor $d\gamma^\beta/dt$ is
  Lorentz-invariant in isolation.  It is most naturally through
  Theorem~\ref{thm:equivalence}---whose proof rests on the
  substitution formula---that we can deduce that the combination
  $p^\alpha\,(d\gamma^\beta/dt)\,\delta^{(3)}$ transforms as a
  tensor.  The miracle is that the Jacobian factors conspire to
  produce a Lorentz-invariant measure along the worldline; the
  change of variables explains \emph{why} they conspire.
\end{remark}

% ──────────────────────────────────────────────────────────────
\subsection{Distributional conservation law}\label{sec:conservation-add}

We now prove the conservation law entirely within the
distributional framework, avoiding the somewhat formal manipulation
of derivatives of delta functions that appears in
Weinberg~\cite[\S2.8]{Weinberg72}.

\begin{theorem}[Conservation of $T^{\alpha\beta}$]%
\label{thm:conservation}
  If each particle is free ($\dd p_j^\alpha/\dd\tau = 0$), then
  \begin{equation}\label{eq:conservation}
    \partial_\beta\, T^{\alpha\beta} = 0
  \end{equation}
  in the distributional sense.  More generally, in the presence of
  external forces,
  \begin{equation}\label{eq:conservation-force}
    \partial_\beta\, T^{\alpha\beta}(x) = G^\alpha(x)\,,
  \end{equation}
  where the \textbf{force density}
  \[
    G^\alpha(x) = \sum_j \int\!\dd\tau\;
      \frac{\dd p_j^\alpha}{\dd\tau}(\tau)\,
      \delta^{(4)}\!\bigl(x - \gamma_j(\tau)\bigr)
  \]
  is the distributional pushforward of the equations of motion.
\end{theorem}

\begin{proof}
  We prove~\eqref{eq:conservation};
  equation~\eqref{eq:conservation-force} follows by the same
  argument, now retaining the nonzero right-hand side
  of~\eqref{eq:cons-ibp} when $\dd p_j^\alpha/\dd\tau \neq 0$.
  For any $\phi \in \mathcal{D}(\Rn{4})$, the standard definition
  of the distributional derivative (integration by parts with a
  sign) gives:
  \begin{align}
    \langle \partial_\beta T^{\alpha\beta},\,\phi\rangle
      &= -\langle T^{\alpha\beta},\,\partial_\beta\phi\rangle
      \notag\\
      &= -\sum_j \int_{\R} p_j^\alpha(\tau)\,
        \dot\gamma_j^\beta(\tau)\,
        \partial_\beta\phi\bigl(\gamma_j(\tau)\bigr)\;\dd\tau
      \notag\\
      &= -\sum_j \int_{\R} p_j^\alpha(\tau)\,
        \frac{\dd}{\dd\tau}\bigl[\phi(\gamma_j(\tau))\bigr]
        \;\dd\tau\,,
        \label{eq:cons-chain}
  \end{align}
  where the last step uses the chain rule
  $\frac{\dd}{\dd\tau}\phi(\gamma(\tau))
  = \dot\gamma^\beta(\tau)\,\partial_\beta\phi(\gamma(\tau))$.
  Since $p_j^\alpha = m_j\dot\gamma_j^\alpha$ and each~$\gamma_j$
  is smooth (Definition~\ref{def:timelike-worldline}), the
  four-momentum $p_j^\alpha(\tau)$ is smooth, so integration by
  parts is justified.  Since $\gamma_j$ is proper
  (Proposition~\ref{prop:Tmunu-well-defined}), the set
  $\gamma_j^{-1}(\operatorname{spt}\phi)$ is compact, so the
  function $\tau \mapsto \phi(\gamma_j(\tau))$ has compact support
  in~$\R$.  The boundary terms therefore vanish, giving:
  \begin{equation}\label{eq:cons-ibp}
    \langle \partial_\beta T^{\alpha\beta},\,\phi\rangle
      = \sum_j \int_{\R}
        \frac{\dd p_j^\alpha}{\dd\tau}(\tau)\,
        \phi\bigl(\gamma_j(\tau)\bigr)\;\dd\tau\,.
  \end{equation}
  For free particles, $\dd p_j^\alpha/\dd\tau = 0$, and the
  right-hand side vanishes for all~$\phi$.
\end{proof}

\begin{remark}
  The elegance of the distributional proof is worth
  emphasising.  In the $3{+}1$ approach, the conservation law
  requires computing
  $\partial_i\bigl[p^\alpha\,(\dd\gamma^\beta/\dd t)\,
  \delta^{(3)}(\vb{x} - \boldsymbol{\gamma}(t))\bigr]
  + \partial_0\bigl[p^\alpha\,(\dd\gamma^0/\dd t)\,
  \delta^{(3)}(\vb{x} - \boldsymbol{\gamma}(t))\bigr]$,
  which involves the multivariable chain rule
  \[
    \frac{\dd}{\dd t}\,\delta^{(3)}\!\bigl(\vb{x}
    - \boldsymbol{\gamma}(t)\bigr)
    = -\frac{\dd\gamma^i}{\dd t}\,
      \frac{\partial}{\partial x^i}\,
      \delta^{(3)}\!\bigl(\vb{x}
      - \boldsymbol{\gamma}(t)\bigr)
  \]
  to convert time derivatives into spatial derivatives.  The
  distributional proof~\eqref{eq:cons-chain}--\eqref{eq:cons-ibp}
  bypasses this entirely: the chain rule on test functions and
  integration by parts do all the work.
\end{remark}

\begin{exercise}\label{ex:conservation-collision}
  Extend Theorem~\ref{thm:conservation} to particles that undergo
  strictly localised collisions (so that worldlines begin and end
  at discrete spacetime events).  Show that $\partial_\beta
  T^{\alpha\beta} = 0$ provided that four-momentum is conserved
  at each collision vertex: $\sum_{n\in c} p_n^\alpha = 0$ for
  each collision~$c$.
  \emph{Hint:} integrate by parts on each worldline segment;
  the boundary terms at each collision vertex sum to zero by
  momentum conservation.
\end{exercise}

% ──────────────────────────────────────────────────────────────
\subsection{Slicing by an arbitrary spacelike hypersurface}%
\label{sec:general-slicing}

The equivalence of the covariant and $3{+}1$ forms does not depend
on the particular choice of time coordinate~$x^0$.  We now state
the general result, which applies to arbitrary spacelike
hypersurfaces and extends naturally to curved spacetimes.

\begin{theorem}[General slicing theorem]%
\label{thm:general-slicing}
  Let $\Sigma = f^{-1}(0)$ be a smooth spacelike Cauchy surface
  in~$(\Rn{4},\eta)$, defined as a level set of a smooth function
  $f\colon\Rn{4}\to\R$ with $\nabla f$ everywhere timelike
  (that is, $\eta^{\alpha\beta}\partial_\alpha f\,\partial_\beta f < 0$).
  Then each worldline~$\gamma_j$ intersects~$\Sigma$
  transversally at a unique point
  $\gamma_j(\tau_j^\ast)$, and the restriction (``slice'') of
  $T^{\alpha\beta}$ to~$\Sigma$ satisfies
  \begin{equation}\label{eq:general-slicing}
    T^{\alpha\beta}\big|_\Sigma(\vb{y})
      = \sum_{j=1}^{N}
        \frac{p_j^\alpha(\tau_j^\ast)\,
              \dot\gamma_j^\beta(\tau_j^\ast)}
             {\bigl|\partial_\gamma f\,
              \dot\gamma_j^\gamma(\tau_j^\ast)\bigr|}\,
        \delta_\Sigma(\vb{y} - \vb{y}_j)\,,
  \end{equation}
  where $\vb{y}$ denotes coordinates on~$\Sigma$, $\vb{y}_j$ is
  the location of the $j$-th crossing on~$\Sigma$, and
  $\delta_\Sigma$ is the delta function with respect to the
  Riemannian measure on~$\Sigma$ induced by the ambient Minkowski
  metric; since $\Sigma$ is spacelike, this induced metric is
  positive definite.  (This coincides with the $3$-dimensional
  Hausdorff measure~$\mathcal{H}^3$ restricted to~$\Sigma$, up
  to identification of~$\Sigma$ with~$\R^3$ via coordinates.)
\end{theorem}

\begin{proof}[Sketch of proof]
  \emph{Transversality.}\enspace
  The covector field $\partial_\alpha f$ satisfies
  $\eta^{\alpha\beta}\partial_\alpha f\,\partial_\beta f < 0$
  (timelike), and the tangent vector $\dot\gamma^\alpha$ is also
  timelike.  By the reverse Cauchy--Schwarz inequality for
  Lorentzian inner products (if $u,v$ are both
  timelike and in the same time-cone, then
  $|g(u,v)| \ge |u|\,|v| > 0$), the contraction
  $\partial_\gamma f \cdot \dot\gamma^\gamma \neq 0$.

  \emph{Uniqueness of intersection.}\enspace
  Consider the composition $h = f\circ\gamma\colon\R\to\R$.
  Its derivative $h'(\tau) = \partial_\gamma f\bigl(\gamma(\tau)\bigr)
  \,\dot\gamma^\gamma(\tau)$ is nonzero everywhere by
  transversality, and has constant sign by continuity (since
  $\R$ is connected).  Hence $h$ is strictly monotone, so
  $h^{-1}(0)$ contains at most one point.  Existence of the
  crossing follows from the Cauchy-surface hypothesis: by
  definition, every inextendible timelike curve meets a Cauchy
  surface, so each complete worldline~$\gamma_j$ intersects
  $\Sigma = f^{-1}(0)$ at least---hence exactly---once.

  \emph{Slicing and the substitution formula.}\enspace
  The proof proceeds in two stages.  First, the classical
  change-of-variables formula applied to
  $h = f\circ\gamma\colon\R\to\R$ converts the proper-time
  integral into an integral with respect to the level-set
  parameter of~$f$, producing the Jacobian factor
  $|\partial_\gamma f \cdot \dot\gamma^\gamma|^{-1}$.
  Second, the identification of the resulting object as a delta
  function on~$\Sigma$ uses the theory of slicing of normal
  $1$-currents (Federer~\cite[Theorem~4.3.2]{Federer69}), which
  provides the natural framework when $\Sigma$ is not a
  coordinate hyperplane.
\end{proof}

\begin{remark}
  For the standard time function $f(x) = x^0 - t_0$, we have
  $\partial_\gamma f = \delta_\gamma^0$, so
  $|\partial_\gamma f\cdot\dot\gamma^\gamma| = |\dot\gamma^0|$.
  Since $\gamma$ is future-directed
  (Definition~\ref{def:timelike-worldline}),
  $|\dot\gamma^0| = d\gamma^0/d\tau$, and the Jacobian factor
  combines with $\dot\gamma^\beta$ to give
  $d\gamma^\beta/dt$---recovering
  Theorem~\ref{thm:equivalence}.  In curved spacetime one
  replaces $f$ with a Cauchy time function and $\delta_\Sigma$
  with the delta function defined by the induced Riemannian metric
  on the Cauchy surface.
\end{remark}

% ──────────────────────────────────────────────────────────────
\subsection{The four-current revisited}\label{sec:four-current-add}

For completeness, we note that the entire development applies
\emph{mutatis mutandis} to the electromagnetic four-current
$J^\alpha$ of \S\ref{sec:four-current}.  Given particles with
charges~$q_j$, define the distributional four-current by
\begin{equation}\label{eq:Jmu-dist}
  \langle J^\alpha,\,\phi\rangle
    = \sum_{j=1}^{N} \int_{\R} q_j\,
      \dot\gamma_j^\alpha(\tau)\,
      \phi\bigl(\gamma_j(\tau)\bigr)\;\dd\tau\,.
\end{equation}
The change-of-variables argument of Theorem~\ref{thm:equivalence} shows that
\begin{equation}\label{eq:Jmu-31}
  J^\alpha(x) = \sum_{j=1}^{N} q_j\,
    \frac{d\gamma_j^\alpha}{dt}(t)\,
    \delta^{(3)}\!\bigl(\vb{x} - \boldsymbol{\gamma}_j(t)\bigr)\,,
\end{equation}
which unifies the charge density~\eqref{eq:charge-density} and
current density~\eqref{eq:current-density}.  The conservation law
$\partial_\alpha J^\alpha = 0$ follows from the same
integration-by-parts argument as
Theorem~\ref{thm:conservation}, using the constancy of charge
along each worldline ($\dd q_j/\dd\tau = 0$).

% ──────────────────────────────────────────────────────────────
\subsection{Discussion}\label{sec:emt-discussion}

\begin{keyresult}[Summary]
  \begin{enumerate}
    \item[\textup{(i)}] The covariant
      form~\eqref{eq:Tmunu-covariant} is the fundamental
      definition of the microscopic energy-momentum tensor: it is a
      well-defined distribution
      (Proposition~\ref{prop:Tmunu-well-defined}) and a Lorentz
      tensor by construction
      (Proposition~\ref{prop:lorentz-cov}).
    \item[\textup{(ii)}] The $3{+}1$ form~\eqref{eq:Tmunu-dt} is
      derived from the covariant form by \emph{change of variables
      along the worldline} (Theorem~\ref{thm:equivalence}).
    \item[\textup{(iii)}] Conservation
      $\partial_\beta T^{\alpha\beta} = 0$ follows from the
      equations of motion by a distributional
      integration-by-parts argument
      (Theorem~\ref{thm:conservation}), without manipulating
      derivatives of delta functions.
    \item[\textup{(iv)}] The construction generalises immediately
      to arbitrary spacelike hypersurfaces via the slicing
      theorem for currents
      (Theorem~\ref{thm:general-slicing}).
  \end{enumerate}
\end{keyresult}

\noindent
Thus Weinberg's beautiful microscopic picture of the
energy-momentum tensor---as the density and flux of four-momentum
carried by point particles---rests on a rigorous distributional
foundation.  The change-of-variables formula along the worldline is the
elementary mechanism that makes the equivalence work; the full
coarea/slicing framework of geometric measure theory becomes
essential for the generalisation to arbitrary Cauchy surfaces.  Recognising this clarifies what would otherwise appear to
be a sleight of hand in the passage from coordinate-time
expressions to manifestly covariant ones.
